Some examples in extended logic counting classes

### 2017-01-06 LICS/extentions.tex
sc

In order to capture classes of optimization functions, we can extend $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$) as follows. 
Given a relational signature $\R$, the set of $\optqso$-formulas over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{multline*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \\ \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{multline*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing (respectively) instead of summing.
\begin{example}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the bigger clique in a graph. Then this can be define in $\optqso$ with the following formula:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
\end{example}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulas constructed from $\qfo$ operators and $\max$ operators (of first- and second-order). The class $\minqso$ is defined analogously changing $\max$ with $\min$. Note that in $\maxqso$ and $\minqso$ the second-order sum and product quantifiers are not allowed. For example, the formulas $\alpha_1$ and $\alpha_6$ are in $\maxqso$ but $\alpha_2$ and $\alpha_3$ are not. As one could expect, $\maxqso$ and $\minqso$ are the logics that we need to capture $\optp$.
\begin{theorem} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ captures $\maxp$ and $\minp$, respectively, over ordered structures.
\end{theorem}
It is important to mention that a similar result was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ and $\minpb$ of problems in $\maxp$ and $\minp$ whose output value is polynomially bounded.
Interestingly, the result is stronger here since our logic has the freedom to use sum and product quantifiers instead of using a max-and-count problem over boolean formulas. 
Finally, it is straightforward to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the first-order quantifier in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.

>>>>

In order to capture classes of optimization functions, we can extend $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$) as follows. 
Given a relational signature $\R$, the set of $\optqso$-formulas over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{multline*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \\ \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{multline*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing (respectively) instead of summing.
\begin{example}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the bigger clique in a graph. Then this can be define in $\optqso$ with the following formula:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
\end{example}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulas constructed from $\qfo$ operators and $\max$ operators (of first- and second-order). The class $\minqso$ is defined analogously changing $\max$ with $\min$. Note that in $\maxqso$ and $\minqso$ the second-order sum and product quantifiers are not allowed. For example, the formulas $\alpha_1$ and $\alpha_6$ are in $\maxqso$ but $\alpha_2$ and $\alpha_3$ are not. As one could expect, $\maxqso$ and $\minqso$ are the logics that we need to capture $\optp$.
\begin{theorem} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ captures $\maxp$ and $\minp$, respectively, over ordered structures.
\end{theorem}
It is important to mention that a similar result was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ and $\minpb$ of problems in $\maxp$ and $\minp$ whose output value is polynomially bounded.
Interestingly, the result is stronger here since our logic has the freedom to use sum and product quantifiers instead of using a max-and-count problem over boolean formulas. 
Finally, it is straightforward to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.
